---
title: "Reproduced analysis of FARS data"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


```{r package and file loading}
setwd("C:/Users/bkim7/R/fars_project/")
library(tidyr)
library(dplyr)
library(ggplot2)
load("data/clean_fars.RData")
source("R/fars_functions.R")

```

#Results

##Percentages of drivers testing positive by drug type, sex, and year group

```{r prevalence of positive drug tests}
clean_fars %>%
mutate(year_cat = cut(year, breaks = c(1999, 2002, 2006, 2010),
labels = c("1999-2002", "2003-2006",
"2007-2010"),
include.lowest = TRUE, right = TRUE)) %>%
filter(!is.na(sex)) %>%
group_by(drug_type, sex, year_cat) %>%
summarize(n_non_missing = sum(!is.na(positive_for_drug)),
positive_test = sum(positive_for_drug, na.rm = TRUE),
perc_positive = round(100 * positive_test / n_non_missing, 1)) %>%
select(drug_type, sex, year_cat, perc_positive) %>%
unite(sex_year_cat, sex, year_cat) %>%
spread(sex_year_cat, perc_positive) %>%
knitr::kable(col.names = c("Drug type", "F 1999-2002",
"F 2003-2006", "F 2007-2010",
"M 1999-2002", "M 2003-2006",
"M 2007-2010"))
```


##Figure 1: Prevalence of nonalcohol drugs in fatally injured drivers by year and age group

```{r figure_1}
#Total percent positive for non-alcohol drugs for each year grouped by agecat
fars_plot1 <- clean_fars %>%
  filter(as.character(drug_type) != "Alcohol") %>%
  filter(!is.na(agecat)) %>%
  group_by(year, agecat, unique_id) %>%
  summarize(positive_test = any(positive_for_drug, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(year, agecat) %>%
  summarize(perc_positive = mean(positive_test) * 100)

ggplot(fars_plot1, aes(x = year, y = perc_positive, shape = agecat)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(limits = c(0, 35)) +
  theme_classic() +
  labs(x = "Year", y = "Positive for Nonalcohol Drugs, %", shape = "Age") 
```

##Figure 2: Prevalence of nonalcohol drugs in fatally injured drivers by year and drug type

```{r figure_2}
fars_plot2 <- clean_fars %>%
  filter(as.character(drug_type) != "Alcohol") %>%
  filter(!is.na(agecat)) %>%
  group_by(year, drug_type) %>%
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test / n_non_missing, 1))

ggplot(fars_plot2, aes(x = year, y = perc_positive, shape = drug_type)) +
  geom_point() +
  geom_line() +
  theme_classic() +
  labs(x = "Year", y = "Positive for Drugs, %", shape = "Drug type") 
```

##Figure 3: Prevalence of cannabinoid drugs in fatally injured drivers by year and age group

```{r figure_3}
fars_plot3 <- clean_fars %>%
  filter(as.character(drug_type) %in% "Cannabinoid",
         !is.na(agecat)) %>%
  group_by(year, agecat) %>%
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test / n_non_missing, 1))

ggplot(fars_plot3, aes(x = year, y = perc_positive, shape = agecat)) +
  geom_point() +
  geom_line() +
  theme_classic() +
  labs(x = "Year", y = "Positive for Cannabinoid, %", shape = "Age") 

```


##Prevalence of drugs in fatally injured drivers for 1999 and 2010 by drugtype

```{r function perc_cis}
table1 <- clean_fars %>%
  filter(year %in% c(1999, 2010)) %>%
  group_by(drug_type, year) %>%
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            percent = perc_cis(x = positive_test, n = n_non_missing)) %>%
  select(drug_type, year, percent) %>%
  spread(key = year, value = percent) %>%
  knitr::kable(col.names = c("Drug Type", "1999", "2010"))
table1
```

##Statistics for testing for trend in prevalence of drugs over study years by drug type using Cochran-Armitage trend test

```{r function test_trend_ca}
drug_list <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
"Stimulant", "Cannabinoid", "Other")
drug_trend_tests_ca <- lapply(drug_list, test_trend_ca)
drug_trend_tests_ca <- dplyr::bind_rows(drug_trend_tests_ca) %>%
dplyr::mutate(drug = drug_list) %>%
dplyr::select(drug, Z, p.value)
drug_trend_tests_ca %>% knitr::kable()
```


##Statistics for testing for trend in prevalence of drugs over study years by drug type using Wald test of logistic regression coefficient for "year"

```{r function test_trend_log_reg}
drug_list <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
"Stimulant", "Cannabinoid", "Other")
drug_trend_tests_log_reg <- lapply(drug_list, test_trend_log_reg)
drug_trend_tests_log_reg <- dplyr::bind_rows(drug_trend_tests_log_reg) %>%
dplyr::mutate(drug = drug_list) %>%
dplyr::select(drug, Z, p.value)
drug_trend_tests_log_reg %>% knitr::kable()
```

